Automatically proving equivalence by type-safe reflection

### 2016-01-18 paper-JFP-DTP/content/6-discussion.tex
Edit introduction
Remaining to do : perfs test & more related work

@@

Coq's ring prover follows the traditional approach of defining functions, and latter on proving many auxiliary lemmas about them. This is particularly adapted to Coq, which has many facilities for the construction of proofs, and an adapted proof mode. However, this approach kind of duplicate the work : they first tell the machine how it works, and latter, why it works. This separation between the world of computations and the world of logic becomes smaller in our approach with a good use of dependent types, that allows to write more specific types, and thus to capture logical properties. The writing of functions can therefore be guided by this more precise type, and this is one of the main benefit of the approach we've followed. Moreover, we almost get the proof of correctness for free, because every little bit of rewriting done for the normalisation of the reflected terms is accompanied by the logic justification which tells why this rewriting can be done --and locally, this justification is always simple !--. In the end, all the proofs we have to do are systematically trivial, since they only contain trivial rewritings with the available hypothesis and the use of the properties of the corresponding algebraic structure. \\

@@

The correctness of the tactic is in fact obtained by construction. Actually, what we really needed was to produce the proof of the equality $a \simeq b$, and not the normalised terms. The normalised term --and more precisely their construction-- was just a support for building the desired proof of $a \simeq b$, which is precisely the proof of correctness. Thus, there isn't any possibility that we've generated a wrong proof as long as we ensure that :
\begin{itemize}
\item We haven't introduced any axioms
\item We haven't introduced any non total\footnote{Contrary to Coq, Idris allows the definition of non-total functions for helping the development of non-finished real-world projects} function or proof used for the building of the main proof $a \simeq b$. Note that the normalization itself could contain some non total functions without breaking the correctness. That would just mean that when using this tactic, the tactic could loop infinitely and the user would never get his answer. That would of course be bad, but that would not produce any inconsistency --as long as none uses these definitions maliciously to prove something false--, because this non total function would be only used in the world of computations and not in the world of logic.
\end{itemize}

Because we've meet these two conditions, we simply can't produce a false proof of a statement, because the typechecker would complain if that would be the case. If Idris internal type theory is consistent, then one can't produce a false lemma without either introducing an axiom nor introducing a non total function and latter using it to create inconsistencies in the logic.

@@

Completeness means that when given an equality which is provable by using the properties of the concerned algebraic structure, the adapted prover should be able to generate a proof of this equality. Having a complete tactic is important, because otherwise the tactic is not always usuable -- and in the worst case it is never usable--, and thus is not really valuable. However, having a formal proof of the completeness in the system itself isn't really interesting. As long as we do not meet an equality that should be provable, and which isn't proved by our tactic, there is no problem. And if this situation does happen, nothing really bad really happened : no inconsistencies have been introduced in the system. The tactic just refuses to do it automatically for the programmer. The user can still attempt it. \\

If we believe that the normal form computed by the reduction function is indeed a normal form --which means that if two terms are equal according to the properties of an algebraic structure, then their normal form should be equal-- then the tactic is complete.
Indeed, 
$meta\_propetry\ :\ \forall\ a\ b,\ a \simeq b \rightarrow norm\ (reflect\ a) = norm\ (reflect\ b)$ is logically equivalent to

$completeness : \ \forall\ a\ b,\ norm\ (reflect\ a) \neq norm(reflect\ b) \rightarrow a \not{\simeq} b$, which is in fact :
$\ \forall\ a\ b,\ decideEq\ (reflect\ a)\ (reflect\ b) = Nothing \rightarrow a \not{\simeq} b$.
The equivalence between $meta\_property$ and $completeness$ can only be done in the meta theory because it is a proof by contradiction, which is equivalent to the law of excluded middle, that we do not have in constructive logics, on which Idris is based.

Question : Does $completeness$ is provable in the system ? Would it be doable for the little example with natural numbers for example ? I want to know if it's something improvable in the theory itself (and in this case why, precisely!), or just something really hard ?
Anyway, it has never been done for this problem (not for Coq's ring prover, nor for Agda's ring prover). Can I do it for the little example with natural numbers though ? \\

Another little advantage of our approach, compared to Coq's approach, is that we have the guarantee that we can't lose any totality because of the $reflect$ function. The reason is that the produced reflected term is indexed over the original --concrete-- value. By this fact, we don't need a $reify$ function. If we want to go from the reflected term to the original --concrete-- value, we can simply return the index. Thus, we have the guarantee that $(reflect\ a)$ is always a faithful representation of $a$ with our approach. Coq's approach doesn't have this guarantee, and could potentially lose some completeness if this property is not always true, as mentioned earlier during the comparison with Coq's implementation.

>>>>

\label{sect:discuss}

@@

Coq's ring prover follows the traditional\footnote{By ``conventional" we have in mind the proof engineering style developed in the Coq'Art ~\cite{BertotC04}}  approach of defining functions, and latter on proving many auxiliary lemmas about them. This is particularly adapted to Coq, which has many facilities for the construction of proofs, and an adapted proof mode. However, this approach kind of duplicate the work : they first tell the machine how it works, and latter, why it works. This separation between the world of computations and the world of logic becomes smaller in our approach with a fine use of dependent types, that allows to write more specific types, and thus to capture logical properties. The writing of functions can therefore be guided by this more precise type, and this is one of the main benefit of the approach we've followed. Moreover, we almost get the proof of correctness for free, because every little bit of rewriting done for the normalisation of the reflected terms is accompanied by the logic justification which tells why this rewriting can be done --and locally, this justification is always simple !--. In the end, all the proofs we have to do are systematically straightforward, since they only contain rewritings with the available hypothesis and the use of the properties of the corresponding algebraic structure. \\

@@

The correctness of the tactic is in fact obtained by construction. Actually, what we really needed to produce was the proof of the equality $a \simeq b$, and not the normalised terms. The normalised term --and more precisely their construction-- was just a support for building the desired proof of $a \simeq b$, which is precisely the proof of correctness. Thus, there isn't any possibility that we've generated a wrong proof as long as we ensure that :
\begin{itemize}
\item We haven't introduced any axioms
\item We haven't introduced any non total\footnote{Contrary to Coq, Idris allows the definition of non-total functions for helping the development of non-finished real-world projects} function or proof used for the building of the main proof $a \simeq b$. Note that the normalization itself could contain some non total functions without breaking the correctness. That would just mean that when using this tactic, the tactic could potentially loop infinitely and the user would never get his answer. That would of course be bad, but that would not produce any inconsistency --as long as none uses these definitions maliciously to prove something false--, because this non total function would be only used in the world of computations and not in the world of logic.
\end{itemize}

Because we've meet these two conditions, we simply can't produce a false proof of a statement, because the typechecker would complain if that would be the case. If Idris internal type theory is consistent and if Idris' internal implementation follows it (and we believe in both!), then one can't produce a false lemma without either introducing an axiom nor introducing a non total function and latter using it maliciously to create an inconsistency in the logic.

@@

Completeness means that when given an equality which is provable by using the properties of the concerned algebraic structure, the adapted prover should be able to generate a proof of this equality. Having a complete tactic is important, because otherwise the tactic is not always usuable -- and in the worst case it is never usable--, and thus is not really valuable. However, having a formal proof of the completeness in the system itself isn't really interesting. As long as we do not meet an equality that should be provable, and which isn't proved by our tactic, there is no problem. And if this situation does happen one day, nothing really bad happened : no inconsistencies would have been introduced in the system. The tactic would just refuse to do it automatically, but the user could still attempt it. \\

In definitive, we want the tactic to be complete, without requiring a formal proof of it. If we believe that the normal form computed by the reduction function is indeed a normal form --which means that if two terms are equal according to the properties of the algebraic structure, then their normal form should be equal-- then the tactic is complete.
Indeed, 
$isNormalForm\ :\ \forall\ a\ b,\ a \simeq b \rightarrow norm\ (reflect\ a) = norm\ (reflect\ b)$ is logically equivalent to

$completeness : \ \forall\ a\ b,\ norm\ (reflect\ a) \neq norm(reflect\ b) \rightarrow a \not{\simeq} b$, which is in fact :
$\ \forall\ a\ b,\ decideEq\ (reflect\ a)\ (reflect\ b) = Nothing \rightarrow a \not{\simeq} b$.
This equivalence between $isNormalForm$ and $completeness$ can only be done in the meta theory because it is a proof by contradiction, which is equivalent to the law of excluded middle, that we do not have in constructive logics, on which Idris is based.
\\

So far, we haven't encountered any goal that should be provable but which failed to be proven automatically. However, if this situation would appear one day, it would certainly come from a small kind of simplification that would have been forgotten somewhere in the hierarchy, and it would be easy to fix it by adding a sub-function that deals with this specific simplification at the appropriated level. And all the more specialised provers that reuse this prover will also be fixed at the same time thanks to the compositionality of our provers. 
\\

Another little advantage of our approach, compared to Coq's approach, is that we have the guarantee that we can't lose any totality because of the $reflect$ function. The reason is that the produced reflected term is indexed over the original --concrete-- value. By this fact, we don't need a $reify$ function. If we want to go from the reflected term to the original --concrete-- value, we can simply return the index. Thus, unlike Coq, we have the formal guarantee that $(reflect\ a)$ is always a faithful representation of $a$ with our approach, which means that we can't lose any totality because of the $reflect$ function.
